Allocation scores, WIS, and CRPS via decision theory

### 2023-08-24 alloscore_manuscript/methods_apg.tex
continuing to reorient toward AS and auxiliary to CRPS

In this section/report/paper/rambling manifesto, we present a methodological perspective on the use scoring rules for forecast evaluation that we see as well-adapted to our motivating research goal to understand and describe the social utility of infectious disease forecasts. Our approach follows the common practice in decision theory of giving central roles to the foundational concepts of state, action, and consequence. From these primitives, we first derive the familiar measures of quantile score (QS), continuous ranked probability score (CRPS) and their aggregates across forecast targets.  We then introduce new scoring rules, the \emph{specific} and \emph{integrated allocation scores}, as reparametrizations of QS and CRPS aggregates that result when part of our definition of consequence is shifted into our definition of action. Specifically, we will trade the regret a decision maker experiences as a consequence of seeing allocated resources go unused for the requirement that a decision maker choose an action that allocates no more than a certain amount in total. This exchange mirrors the duality in mathematical optimization theory between constraints on "decision variables" and the "Lagrange multiplier variables" which quantify the sensitivity of solutions to the constraints, that is, the regret (or relief) experienced when a change in the constraint leads to a change in the optimal cost.
Our computational methods (see ..) in fact make use of this connection.
Something about Bayes risks...

@@

Broadly speaking, we score a probabilistic forecast $F$ by identifying it with a solution to a \emph{decision problem} and rating the utility of that solution.  The term ``decision problem'' refers to a pre-specified set of permissable actions, possible future states, and resulting consequences of actions under various states.  Accordingly, $F$ will receive a "better" or "worse" score in a particular instance when the consequence is better or worse of a rational decision maker using $F$ (and essentially only $F$) to select an action in the face of uncertainty about a future state.  It should be noted immediately that a better or worse score in a single instance may tell us almost nothing about the intrinsic value of the forecasting method that produced $F$.  But as scores accumulate, we can use their distributions across forecasting instances to make inferences as to whether one forecasting method is better or worse than another in a decision theoretic sense.

\subsection{The ingredients of a decision problem with examples}
The basic example of a future state we are considering is the number of individuals that become sick enough from an infectious disease to be in need of some potentially limited resource such as hospital beds, ventilators, medication, or medical staff.  We write $y$ to refer to a particular state such as the quantity of need in a single location or the quantities across multiple locations or times and $Y$ for a future uncertain state which is a random variable with a set of realizable values $\mathcal{Y}$. The forecast $F$ to be scored is a claim of how $Y$ is (or will be) distributed. We also will refer to states as outcomes, which helps to emphasize their future realization.  But we want to clearly distinguish an outcome from a consequence, which as we discuss below, is the result of taking an action under a certain outcome.

@@

And when actions are assigned to states, there are consequences, which we abstractly identify with an element $c$ of a set $\mathcal{C}$. We have two basic examples in mind of what such a $c$ might be.   First, we consider some amount of the regret mentioned above that resources made available in a place or time are not actually needed at that place or time. We emphasize that the nature of this regret is potentially quite amorphous.  In business contexts (e.g., the "newsvendor problem" from inventory management) this regret is often easily quantifiable as monetary waste, but in a public health context, it could... (A desire to better understand over-allocation regret and its relevance to infectious disease forecast evaluation was actually a primary impetus for the present work.) 
Our second and more concrete example of a consequence is the loss incurred when a sick individual is not able to access a resource because it was not made available to them where and when they needed it.   We emphasize here that while we formally identify such loss as a consequence of a particular action being taken under a particular state, it may or may not be an avoidable loss.  That is, it could just as well result from a factor outside a decision maker's control such as a severe resource constraint as from a choice to withhold a resource, motivated, for example, by a desire to reduce waste.

@@

With a view towards a decision problem defined by these examples of sets $\mathcal{X}$, $\mathcal{Y}$, and $\mathcal{C}$ of actions, outcomes, and consequences, we now give a formal three-step procedure for defining and evaluating a \emph{scoring rule} for a forecast $F$ for a future value $y \in \mathcal{Y}$ of $Y$.

\begin{enumerate}
\item Specify a \emph{loss function} $s(x,y)$ that quantifies the utility or disutility of the consequence $c$ of taking action $x$ under outcome $y$.
\item Map the given probabilistic forecast $F$ for $Y$ to an action $x^F$ which ``solves" the decision problem by minimizing the expected loss $E_{F}[s(x,Y)]$ over all $x \in \mathcal{X}$ under the distribution $F$.  Following \cite{dawid2007geometry}, we call $x^F$ the \emph{Bayes act} for $F$. Note that $x^F$ depends not only on $F$ and $s$, but also on the definition of $\mathcal{X}$. For example, having more available resources usually allows for actions incurring less loss.   
\item Assign $F$ the score $s(x^F,y)$, that is, the numerical quantification via $s$ of the consequence under outcome $y$ of accepting $x^F$ as a solution to the decision problem.
\end{enumerate}{}

Thus, given definitions of $\mathcal{X}$, $\mathcal{Y}$, $\mathcal{C}$, and a loss function $s$, we produce a scoring rule $S(F,y):= s(x^F,y)$ that can be applied to any $F$ for any outcome $y \in \mathcal{Y}$.  We emphasize again a key feature of this formulation: the scoring rule $S$ is defined entirely in terms of what we take to be possible actions, states, and consequences along with a function $s$ that assigns values to the possible consequences.

@@

We begin by imagining a public health worker trying to decide at time $t_0$ what quantity $x$ to procure of a protective resource such as beds, staff, or medicine in anticipation of $y$ new infections that will be observed (and create resource demand) at time $t_1>t_0$. The resource has a unit-cost of $C$, and we assign a loss of $L$ to the consequence of a sick individual not having access to the resource. This specifies a loss function  
\begin{align}
s_{Q,C}(x,y) = Cx + L(y-x)_+ \label{eqn:quantile_loss}
\end{align}
where $u_+ = \max(u,0)$.  Loss functions of this general form are referred to as \emph{piece-wise linear}, distinguishing them from smooth non-linear loss functions such as squared error.
The underlying decision problem here consists of action and outcome sets $\mathcal{X},\mathcal{Y}$ both equal to $\mathbb{R}_+$, the non-negative reals.  (We ignore for now any discreteness of the resource.) The consequence set $\mathcal{C}$ is abstractly a set of over-expenditure regrets and numbers of sick individuals with unmet resource need, but the loss function $s$ maps $\mathcal{C}$ into $\mathbb{R}_+$. We also from now on make the assumption that $C < L$ since otherwise -- at least according to this loss function -- there is no ``cost-effectiveness'' in trying to prevent even a single case of unmet need. This would imply a Bayes act $x^F$ that is identically 0, telling us nothing about the value of $F$.

@@

The decision problem leading to the QS may however be complicated by uncertainty at time $t_0$ about the cost $C$ of the resource. We take such uncertainty as requiring the worker to commit in advance to procurement levels $x_i$ for a range of potential costs $C_i < L$ where $i \in 1, \ldots, m$.  Even though only one of the $x_i$ will end up needing to be procured, there does not seem to be any intrinsically decision theoretic principle by which our loss function should ignore counter-factual but exactly specified consequences of the decision under unrealized costs. If we wanted to commit to a \emph{locality} principle for prediction where losses and scores only depend on predictive densities in a neighborhood of the relevant observation, this attempt would falter, and we would need to frame our methods around the log score and it's higher order relatives. But we do not see that as an appropriate principle in public health resourse allocation where  
probabilistic forecasts can be valuable by pushing decision makers in the "right direction" even when they assign low probability to observed outcomes. We instead adopt a ``sensitivity-to-distance'' principle from which piece-wise linear loss functions and the QS can be shown to naturally emerge. It is this principle that makes the loss under alternative costs seem relevant. \apg{These are obviously incomplete thoughts but I feel like there is something here that I've been grasping for for a lng time.}

We take into account the consequences of actions under a set $\mathcal{C} = \{C_i\}$ of possible costs by expanding the loss function to be a weighted mean of quantile losses under certainty about the cost:
\begin{align}
s_{Q,\mathcal{C}, p_C}(x,y) = \frac{1}{m}\sum_{i=1}^{m}p_i(C_ix_i + L(x_i - y)_+). \label{eqn:sum_CL_loss}
\end{align}
Here, $p_C = \{p_i\}, i = 1,\ldots,m$ is a prior distribution on $\mathcal{C}$ and $x = (x_1,\ldots,x_m)$ is the decision vector corresponding to all possible costs.  $p_C$ would be provided by the decision maker (or perhaps a separate expert), and we emphasize that we are considering it here as part of the loss function $s_{Q,\mathcal{C}, p_C}$ that would be formulated in an effort to adapt forecast evaluation to the specifics of a public health decision problem. In practice, $p_C$ has usually been taken to be uniform, which we will show for a particular $\mathcal{C}$ (given $L$) leads to the WIS familiar in ID forecasting hubs. This might be interpreted as deferring to an uninformative prior.

We can also consider $C$ a continuous random variable with a prior density $f_C$ and accommodate this by letting $m \to \infty$ in \eqref{eqn:sum_CL_loss} 
which (with some mathematical care) converts the weighted mean into an integral
\begin{align}
s_{Q,f_C}(x,y) = \int_{0}^{L} (cx(c) + L(x(c) - y)_+) f_C(c)dc{}.
\end{align}

@@

Similarly, given an outcome $y$, evaluation of the scoring rules $s_{Q,\mathcal{C}, p_C}$ and $s_{Q,f_C}$ on the forecast $F$ amounts to vectorizing the
formula for $S_Q(F,y)$. That is,
\begin{align}
S_{Q,\mathcal{C}, p_C}(F,y) &= \frac{1}{m}\sum_{i=1}^{m}p_i(C_i q_{F,\tau_i} + L(q_{F,\tau_i} - y)_+) \\
S_{Q, f_C}(F,y) &= \int_{0}^{L}(cq_{F,\tau(c)} + L(q_{F,\tau(c)} - y)_+) f_C(c)dc.
\end{align}

>>>>

In this section/report/paper/rambling manifesto, we present a methodological perspective on the use scoring rules for forecast evaluation that we see as well-adapted to our motivating research goal to understand and describe the social utility of infectious disease forecasts. Our approach follows the common practice in decision theory of giving central roles to the foundational concepts of state, action, and result as the basic components of a decision problem. From these primitives, we first derive the familiar measures of quantile score (QS), continuous ranked probability score (CRPS) and their aggregates across forecast targets.  We then introduce new scoring rules, the \emph{specific} and \emph{integrated allocation scores} designed to reflect aspects of the decision problem corresponding to the CRPS that emerge when multiple instances are confronted simultaneously.  Specifically, we will consider not only whether using more or less of a resource leads to a better result in a specific problem, but also whether a limited resource is better allocated to one problem rather than another.
This added concern brings into play the duality in mathematical optimization theory between constraints on ``decision variables" and the ``Lagrange multiplier variables" which quantify the sensitivity of solutions to constraints, that is, the degree to which loosening constraints improves results under the optimal action..
Our computational methods (see ..) in fact make use of this connection.
Something about Bayes risks...

@@

Broadly speaking, we score a probabilistic forecast $F$ by identifying it with a solution to a \emph{decision problem} and rating the utility of that solution.  The term ``decision problem'' refers to a pre-specified set of allowed actions, possible future states, and the possible results of allowed actions under those states.  Accordingly, $F$ will receive a ``better" or ``worse" score in a particular instance when the result is better or worse of a rational decision maker using $F$ (and essentially only $F$) to select an action in the face of uncertainty about a future state.  It should be noted immediately that a better or worse score in a single instance may tell us almost nothing about the intrinsic value of the forecasting method that produced $F$.  But as scores accumulate, we can use their distributions across forecasting instances to make inferences as to whether one forecasting method is better or worse than another in a decision theoretic sense.

\subsection{The ingredients of a decision problem with examples}
The basic example of a future state we are considering is the number of individuals that become sick enough from an infectious disease to be in need of some potentially limited resource such as hospital beds, ventilators, medication, or medical staff.  We write $y$ to refer to a particular state such as the quantity of need in a single location or the quantities across multiple locations or times and $Y$ for a future uncertain state which is a random variable with a set of realizable values $\mathcal{Y}$. The forecast $F$ to be scored is a claim of how $Y$ is (or will be) distributed. We also will refer to states as outcomes, which helps to emphasize their future realization.  But we want to clearly distinguish an outcome from a result, which as we discuss below, which is determined by an action under a certain outcome.

@@

And when actions are assigned to states, there are results, which we abstractly identify with an element $r$ of a set $\mathcal{R}$. We have two basic examples in mind of what such a $r$ might be.   First, we consider some amount of the regret mentioned above that resources made available in a place or time are not actually needed at that place or time. We emphasize that the nature of this regret is potentially quite amorphous.  In business contexts (e.g., the ``newsvendor problem" from inventory management) this regret is often easily quantifiable as monetary waste, but in a public health context, it could... (A desire to better understand over-allocation regret and its relevance to infectious disease forecast evaluation was actually a primary impetus for the present work.) 
Our second and more concrete example of a result is the loss incurred when a sick individual is not able to access a resource because it was not made available to them where and when they needed it.   We emphasize here that while we formally identify such loss as a result of a particular action being taken under a particular state, it may or may not be an avoidable loss.  That is, it could just as well result from a factor outside a decision maker's control such as a severe resource constraint as from a choice to withhold a resource, motivated, for example, by a desire to reduce waste.

@@

With a view towards a decision problem defined by these examples of sets $\mathcal{X}$, $\mathcal{Y}$, and $\mathcal{R}$ of actions, outcomes, and results, we now give a formal three-step procedure for defining and evaluating a \emph{scoring rule} for a forecast $F$ for a future value $y \in \mathcal{Y}$ of $Y$.

\begin{enumerate}
\item Specify a \emph{loss function} $s(x,y)$ that quantifies the utility or disutility of the result $r$ of taking action $x$ under outcome $y$.
\item Map the given probabilistic forecast $F$ for $Y$ to an action $x^F$ which ``solves" the decision problem by minimizing the expected loss $E_{F}[s(x,Y)]$ over all $x \in \mathcal{X}$ under the distribution $F$.  Following \cite{dawid2007geometry}, we call $x^F$ the \emph{Bayes act} for $F$. Note that $x^F$ depends not only on $F$ and $s$, but also on the definition of $\mathcal{X}$. For example, having more available resources usually allows for actions incurring less loss.   
\item Assign $F$ the score $s(x^F,y)$, that is, the numerical quantification via $s$ of the result under outcome $y$ of accepting $x^F$ as a solution to the decision problem.
\end{enumerate}{}

Thus, given definitions of $\mathcal{X}$, $\mathcal{Y}$, $\mathcal{C}$, and a loss function $s$, we produce a scoring rule $S(F,y):= s(x^F,y)$ that can be applied to any $F$ for any outcome $y \in \mathcal{Y}$.  We emphasize again a key feature of this formulation: the scoring rule $S$ is defined entirely in terms of what we take to be possible actions, states, and results along with a function $s$ that assigns values to the possible results.

@@

We begin by imagining a public health worker trying to decide at time $t_0$ what quantity $x$ to procure of a protective resource such as beds, staff, or medicine in anticipation of $y$ new infections that will be observed (and create resource demand) at time $t_1>t_0$. The resource has a unit-cost of $C$, and we assign a loss of $L$ to the result that a sick individual cannot access the resource. This specifies a loss function  
\begin{align}
s_{Q,C}(x,y) = Cx + L(y-x)_+ \label{eqn:quantile_loss}
\end{align}
where $u_+ = \max(u,0)$.  Loss functions of this general form are referred to as \emph{piece-wise linear}, distinguishing them from smooth non-linear loss functions such as squared error.
The underlying decision problem here consists of action and outcome sets $\mathcal{X},\mathcal{Y}$ both equal to $\mathbb{R}_+$, the non-negative reals.  (We ignore for now any discreteness of the resource.) The results set $\mathcal{R}$ is abstractly a set of over-expenditure regrets and numbers of sick individuals with unmet resource need, but the loss function $s$ maps $\mathcal{C}$ into $\mathbb{R}_+$. We also from now on make the assumption that $C < L$ since otherwise -- at least according to this loss function -- there is no ``cost-effectiveness'' in trying to prevent even a single case of unmet need. This would imply a Bayes act $x^F$ that is identically 0, telling us nothing about the value of $F$.

@@

The decision problem leading to the QS may however be complicated by uncertainty at time $t_0$ about the cost $C$ of the resource. We take such uncertainty as requiring the worker to commit in advance to procurement levels $x_i$ for a range of potential costs $C_i < L$ where $i \in 1, \ldots, m$.  Even though only one of the $x_i$ will end up needing to be procured, there does not seem to be any intrinsically decision theoretic principle by which our loss function should ignore counter-factual but exactly specified results of the decision under unrealized costs. If we wanted to commit to a \emph{locality} principle for prediction where losses and scores only depend on predictive densities in a neighborhood of the relevant observation, this attempt would falter, and we would need to frame our methods around the log score and it's higher order relatives. But we do not see that as an appropriate principle in public health resourse allocation where  
probabilistic forecasts can be valuable by pushing decision makers in the "right direction" even when they assign low probability to observed outcomes. We instead adopt a ``sensitivity-to-distance'' principle from which piece-wise linear loss functions and the QS can be shown to naturally emerge. It is this principle that makes the loss under alternative costs seem relevant. \apg{These are obviously incomplete thoughts but I feel like there is something here that I've been grasping for for a lng time.}

We take into account the results of actions under a set $\mathcal{C} = \{C_i\}$ of possible costs by expanding the loss function to be a weighted mean of quantile losses under certainty about the cost:
\begin{align}
s_{Q,\mathcal{C}, p_C}(x,y) = \frac{1}{m}\sum_{i=1}^{m}p_i(C_ix_i + L(x_i - y)_+). \label{eqn:sum_CL_loss}
\end{align}
Here, $p_C = \{p_i\}, i = 1,\ldots,m$ is a prior distribution on $\mathcal{C}$ and $x = (x_1,\ldots,x_m)$ is the decision vector corresponding to all possible costs.  $p_C$ would be provided by the decision maker (or perhaps a separate expert), and we emphasize that we are considering it here as part of the loss function $s_{Q,\mathcal{C}, p_C}$ that would be formulated in an effort to adapt forecast evaluation to the specifics of a public health decision problem. In practice, $p_C$ has usually been taken to be uniform, which we will show for a particular $\mathcal{C}$ (given $L$) leads to the WIS familiar in ID forecasting hubs. This might be interpreted as deferring to an uninformative prior.

We can also consider $C$ a continuous random variable with a prior density $f_C$ and accommodate this by letting $m \to \infty$ in \eqref{eqn:sum_CL_loss} 
which (with some mathematical care) converts the weighted mean into an integral
\begin{align}
s_{Q,f_C}(x,y) = \frac{1}{L}\int_{0}^{L} (cx(c) + L(x(c) - y)_+) f_C(c)dc{}.
\end{align}

@@

Similarly, given an outcome $y$, evaluation of the scoring rules $s_{Q,\mathcal{C}, p_C}$ and $s_{Q,f_C}$ on the forecast $F$ amounts to vectorizing the
formula for $S_Q(F,y)$. That is,
\begin{align}
S_{Q,\mathcal{C}, p_C}(F,y) &= \frac{1}{m}\sum_{i=1}^{m}p_i(C_i q_{F,\tau_i} + L(q_{F,\tau_i} - y)_+) \\
S_{Q, f_C}(F,y) &= \frac{1}{L}\int_{0}^{L}(cq_{F,\tau(c)} + L(q_{F,\tau(c)} - y)_+) f_C(c)dc. \label{eqn:CRPS_cform}
\end{align}

How do the WIS and CRPS arise from this construction? Defining $\tau = 1-C/L$ as a random variable gives it the density 
$f_{\tau}(\tau) = -f_C\left(L(1-\tau)\right)/L$, 
and \eqref{eqn:CRPS_cform} becomes 
\begin{align}
S_{Q, f_C}(F,y) &= \int_{0}^{1}((1-\tau)q_{F,\tau} + (y-q_{F,\tau})_{+})f_{\tau}(\tau)d\tau \\
&= \int_{0}^{1}(\mathbf{1}\{y \leq q_{F,\tau}\}-\tau)(q_{F,\tau} - y)f_{\tau}(\tau)d\tau + \int_{0}^{1}(1-\tau)y f_{\tau}(\tau)d\tau \\
&= wCRPS(F,y) + (1-E[\tau])y
\end{align}
where $w$CRPS is the weighted CRPS of \cite{gneiting2011weightedScoringRules} with quantile weight function $\nu(\tau) = \frac{1}{2}f_{\tau}(\tau)$. 
(The $(1-E[\tau])y$ that results from our loss function $s_Q$ adding a $Cy$ term to the standard quantile loss function does not effect the dependence of the scoring rule of $F$.)

Writing $s_{Q,\tau}(x,y) = s_{Q,C}(x,y)/L = (1-\tau)x + (y-x)_{+}$, we have
\begin{align}
s_{Q,\tau}(q_{F,\tau},y) + s_{Q,1-\tau}(q_{F,1-\tau},y) &= q_{F,\tau} + \tau(q_{F,1-\tau} - q_{F,\tau}) + (y - q_{F,\tau})_{+} + (y - q_{F,1-\tau})_{+} \\
&= y + \tau\left(q_{F,1-\tau} - q_{F,\tau} + \frac{1}{\tau}(q_{F,\tau} - y)_{+} + \frac{1}{\tau}(y - q_{F,1-\tau})_{+}\right) \\
&= y + \tau IS_{2\tau}(F,y)
\end{align}
where $IS_{\alpha}(F,y)$ is the \emph{interval score} suggested by expression (43) in \cite{gneiting2007strictly} and formalized by \cite{bracher2021evaluating}.

So when $\mathcal{C} = \mathcal{C}_1\cup (L-\mathcal{C}_1)$ and $p_\mathcal{C}$ is uniform, we get
